# Importing data from multiple files {#multiple-import}

- Dale Barr (October 30, 2019)

Sometimes you have multiple files, and you want to read them into one big table for analysis. The following code allows you to read in a whole bunch of files from a subdirectory.

We'll call the subdirectory where these files live `datadir`; your subdirectory may have a different name.  If the files are not in a subdirectory, but in the same directory as your script, replace `datadir` with a full stop, i.e., `dir(".", "\\.[Cc][Ss][Vv]$")`.

The main work is accomplished using the `map()` function below from the `{purrr}` package, which is part of the tidyverse. The function `map()` applies a function (in this case, `read_csv()`) to each of the elements in the vector supplied as the first argument.

If you want to run the example to see how things work, you'll need to set up your environment with some sample files.

`r hide("Click here to see how to set things up so you can run the example code below")`

```{r setup-dir, echo=TRUE, message=FALSE}
library("tidyverse")

dir.create("datadir", FALSE)

## create fake data files for three subjects
write_csv(iris %>% slice(1:5), file.path("datadir", "S01.csv"))
write_csv(iris %>% slice(6:10), file.path("datadir", "S02.csv"))
write_csv(iris %>% slice(11:15), file.path("datadir", "S03.csv"))
```

`r unhide()`

<br/><br/>

```{r read-them-all}
library("tidyverse")

# "\\.csv$" = find all files ending with csv or CSV
todo <- tibble(filename = dir("datadir", "\\.[Cc][Ss][Vv]$", full.names = TRUE))

all_data <- todo %>%
  mutate(imported = map(filename, read_csv, col_types = "ddddc")) %>%
  unnest(cols = c(imported))

all_data
```

In the above example, the `filename` variable of `todo` contains the values over which the function `read_csv()` is to be applied. The argument `col_types` which we supplied to `map()` is intended to be passed along to `read_csv()`. Any additional arguments that you want to pass along to `read_csv()` can be placed here (e.g., `skip = 1`).

If there is preprocessing you need to do on each file before reading it in, you can write your own function and call that in place of `read_csv()`.

Sometimes the filename contains metadata (e.g., a string identifying the subject: `S01`, `S02` etc) and you want to pull that out. You can do this by creating a new variable using `mutate()` on the filename. In the example below, we strip away the path from the filename (`datadir/S01.csv` to `S01.csv`) using `basename()` and then remove the file extension `.csv` using `tools::file_path_sans_ext()`.

```{r mutate}
all_data_id <- all_data %>%
  mutate(subj_id = basename(filename) %>% tools::file_path_sans_ext()) %>%
  select(subj_id, Sepal.Length:Species)

all_data_id
```

```{r cleanup, include=FALSE}
unlink("datadir", TRUE, TRUE)
```
